Do LSTMs Understand the Licensing of Negative Polarity Items?

### 2021-12-24 paper/paper.tex
Updates

@@

There is, though, little research investigating the feature of licensing of negative polarity items in language. Negative polarity items are a class of words that are only permitted to appear in certain ``negative'' contexts. For example, in English, negative polarity items include words like \textit{ever}, \textit{any}, and \textit{even}.
\begin{exe}
\ex \begin{xlist}
\ex \label{ex:example1} The boy did not ever give up.
\ex[*]{\label{ex:example2} The boy did ever give up.}
\ex[*]{\label{ex:example3} The boy who did not win did ever give up.}
\end{xlist}
\end{exe}
Considering Examples \ref{ex:example1} and \ref{ex:example2}, we note that the word \textit{ever} is only permitted to occur in a sentence when it is licensed by a word such as \textit{nobody} which provides a ``negative'' context. For this basic example, the relationhip between licensor and NPI would not be a particularly challenging task for a language model to learn; if it doesn't see any negative words in environment, then it won't predict an NPI.

However, considering Example \ref{ex:example3}, we see that this simple heuristic fails. Despite the fact that the word \textit{not} preceeds the NPI, the sentence is still ungrammatical. This is because the relationship between the licensor and the negative polarity item is not based on linear word order but rather on the hierarchical structure of a sentence. In particular, for an NPI to be permissible, it must be c-commanded by a licensor. This c-command relation means that the NPI must be a sister of the licensor or a direct descendant of a sister node of the licensor \cite{ladusaw1980notion}.

@@

@@

\subsection{Surprisal}

We assess the ability of an LSTM to predict the distribution of negative polarity items by using the surprisal metric, as applied in Wilcox et. al. \shortcite{wilcox-etal-2018-rnn}. The metric is defined to be the log inverse probability:
\[ S(x_i) = -\log_2 P(x_i \, | \, h_{i-1}), \]
where $x_i$ is the current word, $h_{i-1}$ is the LSTM's hidden state before $x$ and the probability is given by the softmax activation. The surprisal is measured in bits because we take a base two logarithm.

@@

@@

Our general approach to studying these properties of the dependency between NPI and licensor is as follows. We construct a number of test items containing negative polarity items, including variation in whichever independent variable we are curious about (e.g. whether or not the licensor c-commands the NPI). For each item, we construct two variants: one with a licensor, and one without a licensor. We define the licensing interaction to be the difference in surprisals between these two variants:
\[ \textrm{interaction} = S_{\textrm{no license}} - S_{\textrm{license}}. \]
A large licensing interaction suggests that, in a given test item, the presence of a licensor does dampen the model's surprisal at encountering an NPI, while a small licensing interactoin suggests that, in a given test item, the licensor does not dampen the surprisal.

@@

As discussed in the introduction, negative polarity items are only grammatically allowable in negative contexts. The following examples demonstrate this fact because only the sentences with negative contexts (a,c) are grammatically permissible:
\begin{exe}
\ex\label{ex:license-1}
\begin{xlist}
\ex Bill is not here yet.
\ex[*]{Bill is here yet.}
\end{xlist}
\ex\label{ex:license-2}
\begin{xlist}
\ex They hired somebody without any talent.
\ex[*]{They hired somebody with any talent.}
\end{xlist}
\end{exe}
In this subsection, we confirm the hypothesis that surprisal metrics for NPIs are reduced in the presence of an appropriate licensor. As described above, we created 20 test items containing negative polarity items, and for each item, we constructed one variant with a licensor, and one without a licensor, as exemplified in Examples \ref{ex:license-1} and \ref{ex:license-2}.

@@

\begin{figure}
    \centering
    \subfloat[Grammatical]{{\includegraphics[width=.42\linewidth]{good1} }}
    \subfloat[Ungrammatical]{{\includegraphics[width=.42\linewidth]{bad1} }}
    \qquad
    \subfloat[Grammatical]{{\includegraphics[width=.42\linewidth]{good2} }}
    \subfloat[Ungrammatical]{{\includegraphics[width=.42\linewidth]{bad2} }}
    \caption{Plots of surprisal scores for words in the example sentences. Note that plots (a) and (c) are both contextually negative and plots (b) and (d) are positive. As expected, the surprisal score for the NPIs (\textit{yet} in plots (a) and (b) and \textit{any} in plots (c) and (d)) is much higher in positive contexts.}
    \label{fig:surprisal-basic-licensing}
\end{figure}

@@

Measuring instantaneous surprisal upon encountering an NPI, we found a mean licensing interaction of 0.87, which we deemed significant ($t(19 = 1.82$, $p = 0.042$), so we reject the null hypothesis, concluding that the sentences containing a licensor had significantly lower suprisal than those without a licensor.

@@

Overall, this experiment shows that our LSTM model has at least a basic understanding of when negative polarity items should be expected.

@@

As discussed in the the Introduction, there are structural restrictions on the NPI licensing relationship. For an NPI construction to be grammatical, it is not sufficient that a negative licensor precede the NPI in terms of linear word order. In fact, the licensor must c-command the NPI in order for the construction to be grammatical.

@@

In this subsection we attempt to determine whether LSTMs are capable of understanding syntactic structure with regards to negative polarity items. In order to do this we designed 15 test items, each containing an embedded clause and an NPI. For each item, we generated four variants, as exemplified by those in Example \ref{ex:c-command}. If the model has learned that NPIs are only gramatical when they are c-commanded by a licensor, we should expect that the presence of a licensor in the main clause has a significant impact on surprisal, while the presence of a licensor in the embedded clause should not have a significant impact on surpisal. The results are visualized in Figure \ref{fig:c-command-box-plots}.

@@

These results tell us two things. First, they support the result of section \ref{sec:basic} that the presence of a licensor in the main clause reduces surprisal. Second, because there is a significant effect of the embedded licensor, we conclude that the model has not learned to generalize based on underlying sentence structure.

\subsection{Robustness to Intervening Material}
\label{sec:intervening}
Because the relation between licensor and NPI is based on the c-command relation rather than on linear word order, it is not sensitive to the amount of intervening material which occurs between the licensor and the NPI. In Example \ref{ex:intervening}, the NPI \textit{anymore} is licenced by the word \textit{not} in each sentence. The intervening prepositional phrases and relative clauses have no impact on the grammaticality of the sentences.
\begin{exe}
\ex \label{ex:intervening}
\begin{xlist}
\ex We do not go to the supermarket anymore.
\ex We do not go to the supermarket that is owned by Mark anymore.
\ex We do not go down the block to the supermarket that is owned by Mark anymore. 
\end{xlist}
\end{exe}
We showed in Section \ref{sec:basic} that LSTMs exhibit significantly higher surprisal when encountering an unlicensed NPI than a licensed NPI; however, the analysis in Section \ref{sec:c-command} suggests that the model is using a heuristic based on linear word order rather than taking into account the hierarchical structure of the sentence. Given this, we might predict that the amount of intervening material could have an impact on surprisal.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{intervening_pnt_scatter}
        \caption{Instantaneous licensing interaction}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{intervening_cum_scatter}
        \caption{Cumulative licensing interaction}
    \end{subfigure}
    \caption{Plots of licensing interaction of sentences given the amount of words in between the negative licensor and the NPI}
    \label{fig:intervening1}
\end{figure}
\begin{figure*}
    \centering
    \subfloat[Base Sentence]{{\includegraphics[width=.32\textwidth]{short} }}
    \subfloat[5 Additional Words]{{\includegraphics[width=.32\textwidth]{medium} }}
    \subfloat[8 Additional Words]{{\includegraphics[width=.32\textwidth]{long} }}
    \caption{Plots of surprisal scores for words in the modified sentence. The surprisal score of the negative polarity item \textit{anymore} remains relatively unchanged by the amount of intervening material.}
    \label{fig:intervening2}
\end{figure*}
We also did a slightly different experiment aiming to control for variation in the semantic content of the sentences. Here, we designed 10 sentences base sentences with NPI items. Each of the base sentences comes in three varieties: minimal intervening material (enough to make the sentence grammatical), 3-5 words of additional intervening material, and 6-10 additional words between the negative licensor and the polarity item. In all cases the intervening words modify the object of the base sentence. An example of the results for this section are visualized in Figure \ref{fig:intervening2}.

We once again found that there was a weak positive correlation between amount of intervening material and the licensing interaction. However, as seen in the plots, the scores are relatively similar. This test supports the notation that the model's ability to predict the the distribution of NPIs is not affected by the quantity of intervening material.

% \subsection{Ability to License Multiple Negative Polarity Items}
% In most cases, a negative context can license multiple negative polarity items.
% \begin{exe}
% \ex
% \begin{xlist}
% \ex He didn't want to go with anybody anywhere
% \ex [*]{He did want to go with anybody anywhere} 
% \end{xlist}
% \end{exe} 
\section{Positive Polarity Items}
We briefly turn our attention to positive polarity items (PPIs) which are words that can only occur in ``positive'' contexts. English words such as \textit{some, rather,} and \textit{already} are common examples of PPIs. There is significantly less literature discussing PPIs than NPIs, and often the licensing of a positive context is much less explicit than that of a negative context. We are interested in whether LSTMs will exhibit higher surprisal in unlicensed settings, and if so, does the model understand PPIs to the same extent as NPIs. Establishing a positive context in English is not as explicit as establishing a negative context. Sentences default to the positive setting, and thus there is not an equivalent to \textit{not} which licenses a negative context. We hypothesize that the lack of an obvious licensor might make it more difficult for the model to determine whether a PPI is permissible.

\subsection{Licensing of Positive Polarity Items}
Positive polarity items may only be licensed in positive contexts, as is shown for the PPI \textit{rather} in the following example.
\begin{exe}
\ex \label{ex:positive}
\begin{xlist}
\ex I would rather be there
\ex[*]{I wouldn't rather be there}

\end{xlist}
\end{exe}
Plotted in Figure \ref{fig:ppi-basic-licensing} are graphs surprisal of the sentences in \ref{ex:positive}. One can see that the positive polarity item has a slightly lower surprisal score in the grammatically correct context, but the trend is not as pronounced as what was seen with NPIs.

@@

\begin{figure}
    \centering
    \subfloat[Grammatical]{{\includegraphics[width=.42\linewidth]{img/ppigood.png} }}
    \subfloat[Ungrammatical]{{\includegraphics[width=.42\linewidth]{img/ppibad.png} }}
     \caption{The surprisal score for \textit{rather} is slightly lower in the positive context than in the incorrect negative context}
    \label{fig:ppi-basic-licensing}
\end{figure}

@@

This small experiment suggests that it is more difficult for an LSTM to determine the validity of a PPI than an NPI. This result is not particularly surprising because, unlike NPIs, PPIs lack explicit licensors. Thus, in order to correctly predict the distribution of PPIs, the model would need to represent more complex aspects of language.

@@

In Section \ref{sec:basic}, we showed that the presence of a negative licensor in a sentence caused a significant decrease in the model's surprisal upon encountering an NPI. This suggests that the model has learned that NPIs only occur when a negative licensor is present. However, in Section \ref{sec:c-command}, we saw that the model does not distinguish between licensors which c-command the NPI and those that do not. This suggests that, rather than representing the hierarchical structure of the sentence, the model approximates the NPI/licensor relation using a representation based on linear word order and precedence. Finally, in Section \ref{sec:intervening}, we saw that the model's ability to predict the distribution of NPIs was independent of the amount of intervening material.

Despite the fact that the experiments in Section \ref{sec:c-command} suggested that the model does not represent the c-command relation, the results do not differ too dramatically from what is observed in psycholinguistic studies of NPI licensing. As we would expect, listeners with no time constraints tend to judge sentences like those in Examples \ref{ex:c-command-np} and \ref{ex:c-command-pp} as being ungrammatical. However, when fast time constraints are placed on the judgement task, some speakers mistakenly accept sentences like Example \ref{ex:c-command-np} (in which there exists a negative licensor, but it is contained in an embedded clause) as grammatical \cite{xiang2009illusory,xiang2013dependency,parker2016negative}. This suggests that, under time pressure, the human brain might process the NPI licensing relation using a heuristic similar to the one that our model has learned, which takes into account only precedence and not c-command.

These results point towards a number of questions which could be addressed in future research. First of all, we might like to know how well these results generalize to other models. Our homegrown model was trained on a relatively small data set using limited resources; perhaps a larger LSTM like that presented in Jozefowicz et. al. \shortcite{jozefowicz2016exploring} would have a better chance of correctly modeling the c-command relationship. We could also apply a similar analysis to other types of network architectures, such as transformers. Large transformers like the GPT-2 model presented by Radford et. al. \shortcite{radford2019language} and the BERT model introduced in Devlin et. al. \shortcite{devlin2018bert} have both performed very well on a very general set of language modeling tasks. Repeating a similar analysis on GPT-2 or Bert could provide interesting insights into the extent to which transformer models represent hierarchical structure.

>>>>

\usepackage[linguistics]{forest}

@@

There is, though, little research investigating the feature of licensing of negative polarity items in language. Negative polarity items are a class of words that are only permitted to appear in certain ``negative'' contexts. For example, in English, negative polarity items include words like \textit{ever}, \textit{any}, and \textit{even}.
\begin{exe}
\ex \begin{xlist}
\ex \label{ex:example1} The boy did not ever give up.
\ex[*]{\label{ex:example2} The boy did ever give up.}
\ex[*]{\label{ex:example3} The boy who did not win did ever give up.}
\end{xlist}
\end{exe}
Considering Examples \ref{ex:example1} and \ref{ex:example2}, we note that the word \textit{ever} is only permitted to occur in a sentence when it is licensed by a word such as \textit{not} which provides a ``negative'' context. For this basic example, the relationship between licensor and NPI would not be a particularly challenging task for a language model to learn; if it doesn't see any negative words in environment, then it won't predict an NPI.

However, considering Example \ref{ex:example3}, we see that this simple heuristic fails. Despite the fact that the word \textit{not} preceeds the NPI, the sentence is still ungrammatical. This is because the relationship between the licensor and the negative polarity item is not based on linear word order but rather on the hierarchical structure of a sentence. In particular, for an NPI to be permissible, it must be c-commanded by a licensor. This c-command relation means that the NPI must be a sister of the licensor or a direct descendant of a sister node of the licensor \cite{ladusaw1980notion}. Considering the tree below, it is clear that the licensor \textit{not} fails to c-command the NPI \textit{ever}, so the sentence is ungrammatical.

\begin{exe}
\ex \begin{forest}
[TP [NP [The] [N' [N [boy]] [CP [C [who]] [TP [T [did] [not,circle,fill=lightgray]] [VP [win]]]]]] [T' [T [did] [ever,draw,red]] [VP [give up]]]]
\end{forest}
\end{exe}

@@

Jumelet and Hupkes \shortcite{jumelet2018language} investigated neural language models for NPIs via analysis utilizing the \textit{perplexity} metric. This metric is a measure of how well a language model predicts a certain test item; the lower the perplexity score of a sentence is, the better a model was able to predict its tokens. Jumelet et al. \shortcite{jumelet2021language} similarly investigated neural language models for NPIs via analysis utilizing conditional probability distributions and a novel method ranking their model's decoder weights for all tokens based on their similarity with monotonicity weights in their sentence data sets. This is the general extent of the existing literature on NPIs, and neither study makes reference to the surprisal metric, which we introduce in Subsection \ref{sec:surpisal} and contend is a superior metric for analysis.

@@

\subsection{Surprisal}
\label{sec:surpisal}

We assess the ability of an LSTM to predict the distribution of negative polarity items by using the surprisal metric, as applied in Wilcox et. al. \shortcite{wilcox-etal-2018-rnn}. The metric is defined to be the log inverse probability:
\[ S(x_i) = -\log_2 P(x_i \, | \, h_{i-1}), \]
where $x_i$ is the current word, $h_{i-1}$ is the LSTM's hidden state prior to $x_i$ and the probability is given by the softmax activation. The surprisal is measured in bits because we take a base two logarithm.

@@

Surprisal is closely related to the perplexity metric used by Jumelet et. al. \shortcite{jumelet2018language} to evaluate an LSTM's ability to predict NPI distribution. While surpisal can be thought of as a measure of how unexpected a given word is in a sentence, perplexity measures how unexpected the entire sentence is. By measuring surprisal on specific words, or on specific parts of sentences, we are able to achieve a finer-grained analysis than would be possible using only perplexity.

@@

Our general approach to studying these properties of the dependency between NPI and licensor is as follows. We construct a number of test items containing negative polarity items, including variation in whichever independent variable we are curious about (e.g. whether or not the licensor c-commands the NPI). For each item, we construct two variants: one with a licensor, and one without a licensor. We define the licensing interaction to be the difference in surprisals between these two variants:
\[ \textrm{interaction} = S_{\textrm{no license}} - S_{\textrm{license}}. \]
A positive licensing interaction suggests that, in a given test item, the presence of a licensor does suppress the model's surprisal upon encountering an NPI, while a zero (or negative) licensing interaction suggests that, in a given test item, the licensor does not dampen the surprisal. Thus, if LSTMs are able to correctly predict the grammatically of an NPI based on the presence, or lack thereof, of a negative context we would expect to see a statistically significant licencing interaction.

@@

As discussed in the introduction, negative polarity items are only permissible in negative contexts. The following examples demonstrate this fact because only the sentences with negative contexts are grammatically permissible:
\begin{exe}
\ex\label{ex:license-1}
\begin{xlist}
\ex Bill is not here yet.
\ex[*]{Bill is here yet.}
\end{xlist}
\ex\label{ex:license-2}
\begin{xlist}
\ex They hired somebody without any talent.
\ex[*]{They hired somebody with any talent.}
\end{xlist}
\end{exe}
In this subsection, we confirm the hypothesis that surprisal metrics for NPIs are reduced in the presence of an appropriate licensor. As described above, we created 20 test items containing negative polarity items, and for each item, we constructed one variant with a licensor, and one without a licensor, as exemplified in Examples \ref{ex:license-1} and \ref{ex:license-2}.

@@

\begin{figure}
    \centering
    \subfloat[Grammatical]{{\includegraphics[width=.9\linewidth]{good1} }}
    \qquad
    \subfloat[Ungrammatical]{{\includegraphics[width=.9\linewidth]{bad1} }}
    \qquad
    \subfloat[Grammatical]{{\includegraphics[width=.9\linewidth]{good2} }}
    \qquad
    \subfloat[Ungrammatical]{{\includegraphics[width=.9\linewidth]{bad2} }}
    \caption{Plots of surprisal scores for words in the example sentences. Note that plots (a) and (c) are both contextually negative and plots (b) and (d) are positive. As expected, the surprisal score for the NPIs (\textit{yet} in plots (a) and (b) and \textit{any} in plots (c) and (d)) is much higher in positive contexts.}
    \label{fig:surprisal-basic-licensing}
\end{figure}

@@

Measuring instantaneous surprisal upon encountering an NPI, we found a mean licensing interaction of 0.87, which we deemed significant ($t(19) = 1.82$, $p = 0.042$), so we reject the null hypothesis that the instantaneous surprisal of a negative polarity item is unchanged by the sentence context. Therefore, we conclude that sentences containing a negative licensor had significantly lower suprisal than those without a licensor.

@@

We conclude that LSTMs understand that the presence of a negative licensor in a sentence increases the probability that a negative polarity item will be present. Overall, this signifies that our LSTM model has at least a basic understanding of when negative polarity items should be expected.

@@

As discussed in the Introduction, there are structural restrictions on the NPI licensing relationship. For an NPI construction to be grammatical, it is not sufficient that a negative licensor precede the NPI in terms of linear word order. In fact, the licensor must c-command the NPI in order for the construction to be grammatical.

@@

In order to analyze the model's performance on sentences like these, we designed 15 test items, each containing an embedded clause and an NPI. For each item, we generated four variants, as exemplified by those in Example \ref{ex:c-command}. If the model has learned that NPIs are only gramatical when they are c-commanded by a licensor, we should expect that the presence of a licensor in the main clause has a significant impact on surprisal, while the presence of a licensor in the embedded clause should not have a significant impact on surpisal. The results are visualized in Figure \ref{fig:c-command-box-plots}.

@@

These results tell us two things. First, they support the result of Section \ref{sec:basic} that the presence of a licensor in the main clause reduces surprisal. Second, because there is a significant effect of the embedded licensor, we conclude that the model has not learned to generalize based on underlying sentence structure. Instead, it appears that the model incorrectly predicts that the presence of a negative licensor, regardless of its position in a sentence, increases the probability for a negative polarity item to exist.

\subsection{Robustness to Intervening Material}
\label{sec:intervening}
Because the relation between licensor and NPI is based on the c-command relation rather than on linear word order, it is not sensitive to the amount of intervening material which occurs between the licensor and the NPI. In Example \ref{ex:intervening}, the NPI \textit{anymore} is licenced by the word \textit{not} in each sentence. The intervening prepositional phrases and relative clauses have no impact on the grammaticality of the sentences.
\begin{exe}
\ex \label{ex:intervening}
\begin{xlist}
\ex We do not go to the supermarket anymore.
\ex We do not go to the supermarket that is owned by Mark anymore.
\ex We do not go down the block to the supermarket that is owned by Mark anymore. 
\end{xlist}
\end{exe}
We showed in Section \ref{sec:basic} that LSTMs exhibit significantly higher surprisal when encountering an unlicensed NPI than a licensed NPI; however, the analysis in Section \ref{sec:c-command} suggests that the model is using a heuristic based on linear word order rather than taking into account the hierarchical structure of the sentence. Given this, we might predict that the amount of intervening material could have an impact on surprisal because the model could forget about the previous presence of a negative licensor as more words are placed between it and an NPI.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{intervening_pnt_scatter}
        \caption{Instantaneous licensing interaction}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{intervening_cum_scatter}
        \caption{Cumulative licensing interaction}
    \end{subfigure}
    \caption{Plots of licensing interaction of sentences given the amount of words in between the negative licensor and the NPI}
    \label{fig:intervening1}
\end{figure}
\begin{figure*}
    \centering
    \subfloat[Base Sentence]{{\includegraphics[width=.49\textwidth]{short} }}
    \subfloat[8 Additional Words]{{\includegraphics[width=.49\textwidth]{long} }}
    \caption{Plots of surprisal scores for words in the modified sentence. The surprisal score of the negative polarity item \textit{anymore} remains relatively unchanged by the amount of intervening material. We see that the instantaneous surprisal of the NPI  differs by under 14\% in this example.}
    \label{fig:intervening2}
\end{figure*}
We also did a slightly different experiment aiming to control for variation in the semantic content of the sentences. Here, we designed 10 sentences base sentences with NPI items. Each of the base sentences comes in three varieties: minimal intervening material (enough to make the sentence grammatical), 3-5 words of additional intervening material, and 6-10 additional words between the negative licensor and the polarity item. In all cases the intervening words modify the object of the base sentence. An example of the results for this section are visualized in Figure \ref{fig:intervening2}. One can see that the length of intervening material has a marginal effect on the surprisal metric on the negative polarity item.

We once again found that there was a weak positive correlation between amount of intervening material and the licensing interaction. However as seen in Figure \ref{fig:intervening2}, which are a good representation of the overall data, the scores are relatively similar. This test supports the notion that the model's ability to predict the the distribution of NPIs is not significantly affected by the quantity of intervening material.

% \subsection{Ability to License Multiple Negative Polarity Items}
% In most cases, a negative context can license multiple negative polarity items.
% \begin{exe}
% \ex
% \begin{xlist}
% \ex He didn't want to go with anybody anywhere
% \ex [*]{He did want to go with anybody anywhere} 
% \end{xlist}
% \end{exe} 
\section{Positive Polarity Items}
We briefly turn our attention to another class of words that are off interest, positive polarity items (PPIs). PPIs can only occur in ``positive'' contexts. English words such as \textit{some, rather,} and \textit{already} are common examples of PPIs. There is significantly less literature discussing PPIs than NPIs, and often the licensing of a positive context is much less explicit than that of a negative context. We are interested in whether LSTMs will exhibit higher surprisal in unlicensed settings, and if so, does the model understand PPIs to the same extent as NPIs. Establishing a positive context in English is not as explicit as establishing a negative context. Declarative sentences in English default to a positive setting, and thus there is not an equivalent to \textit{not} which licenses a negative context. We hypothesize that the lack of an obvious licensor might make it more difficult for the model to determine whether a PPI is permissible.

\subsection{Licensing of Positive Polarity Items}
\label{sec:ppi}
Positive polarity items may only be licensed in positive contexts, as is shown for the PPIs \textit{rather} and \textit{already} in the following examples.
\begin{exe}
\ex \label{ex:positive}
\begin{xlist}
\ex I would rather be there
\ex[*]{I wouldn't rather be there} 
\end{xlist}
\ex \label{ex:positive2}
\begin{xlist}
\ex Evan is already in jail
\ex[*]{Evan isn't already in jail}

\end{xlist}
\end{exe}
Plotted in Figure \ref{fig:ppi-basic-licensing} are graphs surprisal of the sentences in \ref{ex:positive}. One can see that the positive polarity item has a slightly lower surprisal score in the grammatically correct context, but the trend is not as pronounced as what was seen with NPIs. These plots are a good representation of the general trend that we observed with positive polarity items in different sentence contexts.

@@

\begin{figure}
    \centering
    \subfloat[Grammatical]{{\includegraphics[width=.8\linewidth]{img/ppigood.png} }}
    \qquad
    \subfloat[Ungrammatical]{{\includegraphics[width=.8\linewidth]{img/ppibad.png} }}
     \caption{The surprisal score for \textit{rather} is slightly lower in the positive context than in the incorrect negative context}
    \label{fig:ppi-basic-licensing}
\end{figure}

@@

This small experiment suggests that it is more difficult for an LSTM to determine the validity of a PPI than an NPI. This result is not particularly surprising because, unlike NPIs, PPIs lack explicit licensors. Thus, in order to correctly predict the distribution of PPIs, the model would need to represent more complex aspects of language rather than solely establishing a relationship between \textit{not} and NPIs.

@@

In Section \ref{sec:basic}, we showed that the presence of a negative licensor in a sentence caused a significant decrease in the model's surprisal upon encountering an NPI. This suggests that the model has learned that NPIs only occur when a negative licensor is present. However, in Section \ref{sec:c-command}, we saw that the model does not distinguish between licensors which c-command the NPI and those that do not. This suggests that, rather than representing the hierarchical structure of the sentence, the model approximates the NPI/licensor relation using a representation based on linear word order and precedence. In Section \ref{sec:intervening}, we saw that the model's ability to predict the distribution of NPIs was independent of the amount of intervening material. Finally, in section \ref{sec:ppi} we observed that LSTMs do not assign a statistically significant difference in surprisal to positive polarity items in positive and negative contexts. This provides more evidence that the model merely understands that the words \textit{not} and \textit{no} often precede NPIs because PPIs are syntactically identical to NPIs, but they lack those explicit licensors.

These results point towards a number of questions which could be addressed in future research. First of all, it could be interesting to compare our model's knowledge of the NPI/licensor relation with the knowledge of human subjects. Despite the fact that the experiments in Section \ref{sec:c-command} suggested that the model does not represent the c-command relation, the results do not differ too dramatically from what is observed in psycholinguistic studies of NPI licensing in humans. As we would expect, human subjects with no time constraints tend to judge sentences like those in Examples \ref{ex:c-command-np} and \ref{ex:c-command-pp} as being ungrammatical. However, when strict time constraints are placed on the judgement task, some speakers mistakenly accept sentences like Example \ref{ex:c-command-np} (in which there exists a negative licensor, but it is contained in an embedded clause) as grammatical \cite{xiang2009illusory,xiang2013dependency,parker2016negative}. These constructions which are mistakenly accepted as grammatical are referred to as Negative Polarity Illusions, and their existence suggests that, under time pressure, the human brain might process the NPI licensing relation using a heuristic similar to the one that our model has learned, which takes into account only precedence and not c-command.

In order to analyse this apparent similarity more thoroughly, we could perform an experiment in which the same sentences are presented to both an LSTM language model and to a human subject. It has been observed that, in the general setting, surprisal in state of the art language models is correlated with human sentence processing difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}. By focusing specifically on sentences like those in Example \ref{ex:c-command}, we could do a more controlled comparison of how LSTMs and human subjects respond to Negative Polarity Illusions.

Another route for future research, which would not require the use of human test subjects, might involve studying how well our results generalize to other models. Our homegrown model was trained on a relatively small data set using limited resources; perhaps a larger LSTM like that presented in Jozefowicz et. al. \shortcite{jozefowicz2016exploring} would have a better chance of correctly modeling the c-command relationship. We could also apply a similar analysis to other types of network architectures, such as transformers. Large transformers like the GPT-2 model presented by Radford et. al. \shortcite{radford2019language} and the BERT model introduced in Devlin et. al. \shortcite{devlin2018bert} have both performed very well on a very general set of language modeling tasks, and it has been suggested that they have some ability to represent hierarchical structure \cite{jawahar2019does,clark2019does}. Repeating a similar analysis on GPT-2 or Bert could provide interesting insights into the differences between LSTM models and transformer models, and the ability of these different model types to represent structural data.
